SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-04-01 2017-Xudong-SMT/ActaInformatica.tex
more cosmetic changes

\abstract{Open pNets are formal models that can express the behaviour of open systems, either
	synchronous, asynchronous, or heterogeneous. They are endowed with a symbolic operational semantics in
	terms of open automata, which allows us to check properties of
	such systems in a compositional manner. 
	We present an algorithm computing these semantics,
	building predicates expressing the synchronisation conditions
	between the events of the pNet sub-systems. Checking such
	predicates requires symbolic reasoning about first order logics and application-specific data. We use the Z3 SMT engine to check
	satisfiability of the predicates.
	%As an industrial oriented use-case, %we use so-called ``architectures''
	%for BIP systems, that have been used in the framework of an ESA project and to specify the control software of a nanosatellite at the EPFL Space Engineering Center. 
%	We also present several use-cases, in particular, encoding a BIP architecture
%	extended with explicit data and computing its
%	open automaton semantics. This automaton may be used to prove
%	behavioural properties such as safety and liveness.
We also propose and implement an optimised
	algorithm that performs part of the pruning on-the-fly, and show 
	its correctness with respect to the original one.
	We illustrate the approach using two use-cases: the first one is a
	classical process-algebra operator for which we provide several
	encodings, and prove some basic properties. The second one is  industry-oriented and based on the so-called ``BIP architectures'', which have
	been used to specify the
	control software of a nanosatellite at the EPFL Space Engineering
	Center. We use pNets to encode a BIP architecture
	extended with explicit data, compute its semantics and discuss its
	properties, and then show how our algorithms scale up, using a
	composition of two such architectures.}

@@

	\begin{itemize}
		\item $Sorts$ is a set of data sorts.
		% \noteSB{Commented out a footnote here\mdash check if still necessary.}
		%% \footnote{Later we may want to extend this with Sort constructors,
		%%   like Array or Pair, but this is not needed now}
		\item $\mathit{Constrs}$ is a set of \emph{constructor operators}: for each $\mathit{Con} \in \mathit{Constrs}$, $arity(Con)=n \in \mathbb{N}$ is its arity 
		and '$Con : (\mathit{sel}_1,\mathit{sort}_1), \dots, (\mathit{sel}_n,\mathit{sort}_n) \rightarrow \mathit{sort}$' is its signature with the associated selectors.
		For each argument, the pair $(\mathit{sel}_i,\mathit{sort}_i)$ defines an auxiliary
		operator of name $\mathit{sel}_i$ with signature $\mathit{sel}_i : \mathit{sort} \rightarrow \mathit{sort}_i$.
		\item $Ops$ is a set of \emph{auxiliary operators}, with their
		arity and signature, of the form: $Op : \mathit{sort}_1, \dots,  \mathit{sort}_n \rightarrow
		\mathit{sort}$.
		\item $\mathit{Constrs}(\mathit{sortname})$ and $\mathit{Sels}(\mathit{sortname})$ are, respectively, the sets of
		constructors and selectors of the sort $\mathit{sortname}$.
	\end{itemize}
	%  \noteSB{I think this can be removed}\hl{All sorts and operator names must be distinct.}
	Constructors of arity 0 are called \emph{constants}, and denoted $\mathit{Consts}(\APres)$.
	%  $\mathit{Consts}(\APres) = \{\mathit{Con} \in \mathit{Constrs}\,|\, \mathit{arity}(\mathit{Con})=0\}$.
\end{definition}

@@

\paragraph{States of open pNets.}\label{def-states}
A state of an open pNet is a tuple of the
states of its leaves (in which we denote tuples
in structured states as $\triangleleft\ldots\triangleright$).
For any pNet $p$, let $\overline{\mathit{Leaves}} = \langle S_i,{s_i}_0, \to_i\rangle^{i \in L}$ be the set of PLTSs at its leaves,
then $\mathit{States}(p) = \{\ostate{s_i^{i\in L}}
|\, \forall i\in L. s_i \in S_i\}$.
A PLTS could be its own single leave:
$\mathit{States}(\langle S,s_0, \to\rangle) = \{\ostate{s}|\, s \in S\}$.
The initial state is defined as:
$\mathit{InitState}(p) = \ostate{{{s_i}_0}^{i\in L}}$.

@@

In \cite{henrio:Forte2016} we also proved the following result, which
ensures the termination of our semantic construction algorithm.
\begin{theorem}[Finiteness]
  \label{Th:finiteness}
   Let $pnet=\langle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\rangle$ be an open pNet  with leaves $l_i^{i\in I}$ and holes $h_j^{j\in
 	J}$, where the sets $I$ and $J$ are finite. Suppose the synchronisation vectors of all pNets 
 included in  $pnet$   are finite, and 
 % $\forall i \in L.\, finite{(states(l_i))} \text{ and } l_i$ 
 $l_i $ has a finite number of state variables for each $i\in I$. Then the semantics of $pnet$ is an open automaton 
 %$\mathcal{T}$ 
 with finitely many states and transitions.
\end{theorem}
  Remark that all the elements of such pNets and open automata are symbolic, and they can represent many classes of unbounded systems.

\section{Generation of Open Automata}
\label{section:implementation}
In this section we describe the algorithm implementing the pNet
semantics, its new ``on-the-fly'' version, and the interaction with the
Z3 SMT solver. Under the hypotheses of Theorem \ref{Th:finiteness} 
both algorithms terminate.

@@

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a PLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$;
          add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State $\mathit{L1}$=filterUnsatTransitions($\mathit{L,Assignments}$);
\State $\mathit{OA_0}$=makeReachableSubAutomaton($\mathit{InitState}(P),L1)$;
\Repeat $\mathit{OA_n}$=RefineReachableSubAutomaton($\mathit{OA_{n-1}}$)
\Until  $\mathit{OA_n}=\mathit{OA_{n-1}}$; \Comment Fixpoint reached
\State \Return $\mathit{OA_n}$;

@@

Before the SAT checking step, it may contain symbolic transitions that
do not have valid ground instanciations. This is not logically
incorrect, but is naturally not an optimal representation. So we prune
these UNSAT transitions: at the end of the open automaton generation,
the predicate of each OT is translated 
into SMTlib assertions (see Section~\ref{section:pruning}), and checked for
satisfiability (function \texttt{filterUnsatTransitions}). 
Then we compute the set of reachable transitions and states to build
the final open automaton (more details in Section~\ref{section:BuildingTheOpenAutomaton}).

The inside loop (\emph{MakeTransitions} method) applies recursively
the semantic rules following the structure of the pNet.
When applying a rule to a PLTS at the leaves, we simply take the PLTS transitions of the
corresponding local state and use the semantic rule to build the OT.
When applying a rule to a pNet node we use two methods, combining and
matching, to generate the open transitions in a hierarchical manner,
as shown in Algorithm~\ref{alg2}. This method directly manages the
holes of the node, so \emph{MakeTransitions} is never called on a hole.

@@

\paragraph{Filtering.}
While matching a vector with a combination tuple, \emph{Matching}
tries to filter out some incompatibilities; there may be several
reasons why the matching would fail: 
\begin{itemize}
  \item if some subnet is marked as inactive in the vector, and the
    chosen combination has an active behaviour at this position, or vice-versa;
  \item if some subnet action expression in the vector does not match (by
    pattern-matching) the corresponding action expression in $C$;
  \item or if the whole set of active subnet actions in the vector
    cannot be matched (by unification) with the corresponding action expressions
      in the tuple $C$ (this is strictly stronger than the
      previous filter). 
\end {itemize}

@@

  Conversely, if $|S|_{L1_{reach}}=n+1$, we get a similar chain in
  $L1_{reach}$, with the transition $S_n \xrightarrow{l_{n}}S$.  Then $ S_n \in
  States(L1_{reach})$ and $l_n$ is satisfiable. By induction, we have
  $S_n \in States(L2_{reach})=States(L2_{sat})$, so $l_n$ will have
  been checked ``on-the-fly'', in line 6 of Algorithm~\ref{alg3}. Then
  the transition $S_n \xrightarrow{l_{n}}S$ will be added to $L2_{sat}$, and thus $S \in States(L2_{reach})$.

@@

Here the behavior of the hole ``P'' ($a1$ in Figure~\ref{schema:unsat-ot}) , after matching with the synchronisation vectors, appears as a variable
``|a1:sva\_SVA0:15:1|'', and ``R'' ($acc(x2)$) as ``|(ACT delta
|x:sva\_SV4:1:2|''. Line 16 requires that $a1 = \delta(x2)$ to match
the second synchronisation vector of the root pNet node, and lines
17-18 that $a1$ does not match $\delta(x1)$ for any $x1$. 
We also display  the diagnosis (``sat'' or ``unsat'') generated by Z3.

@@

Figure~\ref{schema:smt-lib} shows the decomposition of the
predicate into a set of asserts, each encoding an elementary equality,
inequality, or a guard. 
The result (sat or unsat) of the final \texttt{check-sat} command in
the translation () is then decoded.

@@

\paragraph{Computed open automaton}

@@

        \medskip
        In the basic cases (EnableState, EnableData) there is no
        benefit, and even an overhead in the case of EnableData,
        because of the extra step of checking SAT at the end of the
        algorithm.
But as soon as there are more OTs in the system and not all
states are reachable, these tests show a significant benefit of the ``smart'' strategy, in
the considered use-cases, with 50\% of time saving for the FailureTimer,
over to 80\% for the composition, depending
        on the nature of the examples, but increasing significantly
        with their complexity.

@@

The formal definitions and properties of the open pNets model were
given in \cite{henrio:Forte2016}. In the current work we describe an implementation of
the model and its semantics construction, including its interaction
with the Z3 SMT engine.
The implementation has two parts: the first is an algorithm
that builds all possible combinations of synchronisations through the
pNet hierarchical structure. The result is a so-called
\emph{open automaton}, whose transitions contain predicates relating
the actions of the pNet holes and controllers. Some of the open
transitions obtained at this step  may
contain predicates which do not represent any possible concrete
instantiations. 
In the second part of the tool we use the SMT solver Z3 to check the
satisfiability of the predicate in each open transition. 
To this end, we encode into Z3 the representations of the action algebra and
the predicates before submitting them to the
Z3 solver. 
%
%\hl{%
%
We have used a running example based on two possible
encodings of the Enable operator of LOTOS. The automata computed with
the algorithm can be used to prove equational properties of the
operator.

>>>>

\abstract{Open pNets are formal models that can express the behaviour of open systems, either
	synchronous, asynchronous, or heterogeneous. They are endowed with a symbolic operational semantics in
	terms of open automata, which allows us to check properties of
	such systems in a compositional manner. 
	We present an algorithm computing these semantics,
	building predicates expressing the synchronisation conditions
	between the events of the pNet sub-systems. Checking such
	predicates requires symbolic reasoning about first order logics and application-specific data. We use the Z3 SMT engine to check
	satisfiability of the predicates.
	%As an industrial oriented use-case, %we use so-called ``architectures''
	%for BIP systems, that have been used in the framework of an ESA project and to specify the control software of a nanosatellite at the EPFL Space Engineering Center. 
%	We also present several use-cases, in particular, encoding a BIP architecture
%	extended with explicit data and computing its
%	open automaton semantics. This automaton may be used to prove
%	behavioural properties such as safety and liveness.
We also propose and implement an optimised
	algorithm that performs part of the pruning on the fly, and show 
	its correctness with respect to the original one.
	We illustrate the approach using two use-cases: the first one is a
	classical process-algebra operator for which we provide several
	encodings, and prove some basic properties. The second one is  industry-oriented and based on the so-called ``BIP architectures'', which have
	been used to specify the
	control software of a nanosatellite at the EPFL Space Engineering
	Center. We use pNets to encode a BIP architecture
	extended with explicit data, compute its semantics and discuss its
	properties, and then show how our algorithms scale up, using a
	composition of two such architectures.}

@@

	\begin{itemize}
		\item $Sorts$ is a set of data sorts.
		% \noteSB{Commented out a footnote here\mdash check if still necessary.}
		%% \footnote{Later we may want to extend this with Sort constructors,
		%%   like Array or Pair, but this is not needed now}
		\item $\mathit{Constrs}$ is a set of \emph{constructor operators}: for each $\mathit{Con} \in \mathit{Constrs}$, $arity(Con)=n \in \mathbb{N}$ is its arity 
		and '$Con : (\mathit{sel}_1,\mathit{sort}_1), \dots, (\mathit{sel}_n,\mathit{sort}_n) \rightarrow \mathit{sort}$' is its signature with the associated selectors.
		For each argument, the pair $(\mathit{sel}_i,\mathit{sort}_i)$ defines an auxiliary
		operator of name $\mathit{sel}_i$ with signature $\mathit{sel}_i : \mathit{sort} \rightarrow \mathit{sort}_i$.
		\item $Ops$ is a set of \emph{auxiliary operators}, with their
		arity and signature, of the form: $Op : \mathit{sort}_1, \dots,  \mathit{sort}_n \rightarrow
		\mathit{sort}$.
		\item $\mathit{Constrs}(\mathit{sortname})$ and $\mathit{Sels}(\mathit{sortname})$ are, respectively, the sets of
		constructors and selectors of the sort $\mathit{sortname}$.
	\end{itemize}
	%  \noteSB{I think this can be removed}\hl{All sorts and operator names must be distinct.}
	Constructors of arity 0 are called \emph{constants}, and denoted by $\mathit{Consts}(\APres)$.
	%  $\mathit{Consts}(\APres) = \{\mathit{Con} \in \mathit{Constrs}\,|\, \mathit{arity}(\mathit{Con})=0\}$.
\end{definition}

@@

\paragraph{States of open pNets.}\label{def-states}
A state of an open pNet is a tuple of the
states of its leaves. We denote tuples
in structured states in the form $\triangleleft\ldots\triangleright$.
For any pNet $p$, let $\langle S_i,{s_i}_0, \to_i\rangle^{i \in L}$ be the set of PLTSs at its leaves,
then $\mathit{States}(p) = \{\ostate{s_i^{i\in L}}
|\, \forall i\in L. s_i \in S_i\}$.
A PLTS could be its own single leave:
$\mathit{States}(\langle S,s_0, \to\rangle) = \{\ostate{s}|\, s \in S\}$.
The initial state is defined as:
$\mathit{InitState}(p) = \ostate{{{s_i}_0}^{i\in L}}$.

@@

In \cite{henrio:Forte2016} we also proved the following result, which
ensures the termination of our semantic construction algorithm.
\begin{theorem}[Finiteness]
  \label{Th:finiteness}
   Let $pnet=\langle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\rangle$ be an open pNet  with leaves $l_i^{i\in I}$ and holes $h_j^{j\in
 	J}$, where the sets $I$ and $J$ are finite. Suppose the synchronisation vectors of all pNets 
 included in  $pnet$   are finite, and 
 % $\forall i \in L.\, finite{(states(l_i))} \text{ and } l_i$ 
 $l_i $ has a finite number of state variables for each $i\in I$. Then the semantics of $pnet$ is an open automaton 
 %$\mathcal{T}$ 
 with finitely many states and transitions.
\end{theorem}
Notice that all the elements of such pNets and open automata are symbolic, and they can represent many classes of unbounded systems.

\section{Generation of Open Automata}
\label{section:implementation}
In this section we describe an algorithm implementing the pNet
semantics, its new ``on-the-fly'' version, and the interaction with the
Z3 SMT solver. Under the hypotheses of Theorem \ref{Th:finiteness} 
both algorithms terminate.

@@

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a PLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$;
          add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State $\mathit{L1}$=filterUnsatTransitions($\mathit{L,Assignments}$);
\State $\mathit{OA_0}$=makeReachableSubAutomaton($\mathit{InitState}(P),L1)$;
\Repeat $\mathit{OA_n}$=RefineReachableSubAutomaton($\mathit{OA_{n-1}}$)
\Until  $\mathit{OA_n}=\mathit{OA_{n-1}}$;   \hfill $\setminus \setminus$ Fixed point reached
\State \Return $\mathit{OA_n}$;

@@

Before the SAT checking step, it may contain symbolic transitions that
do not have valid ground instantiations. This is not logically
incorrect, but is naturally not an optimal representation. So we prune
these UNSAT transitions: at the end of the open automaton generation,
the predicate of each OT is translated 
into SMTlib assertions (see Section~\ref{section:pruning}), and checked for
satisfiability (function \texttt{filterUnsatTransitions}). 
Then we compute the set of reachable transitions and states to build
the final open automaton (more details in Section~\ref{section:BuildingTheOpenAutomaton}).

The inside loop (\emph{MakeTransitions} method) applies recursively
the semantic rules following the structure of the pNet.
When applying a rule to a PLTS at the leaves, we simply take the PLTS transitions of the
corresponding local state and use the semantic rules to build the OT.
When applying a rule to a pNet node we use two methods, \emph{combining} and
\emph{matching}, to generate the open transitions in a hierarchical manner,
as shown in Algorithm~\ref{alg2}. This method directly manages the
holes of the node, so \emph{MakeTransitions} is never called on a hole.

@@

\paragraph{Filtering.}
While matching a vector with a combination tuple, \emph{Matching}
tries to filter out some incompatibilities; there may be several
reasons why the matching would fail: 
\begin{itemize}
  \item if some subnet is marked as inactive in the vector, and the
    chosen combination has an active behaviour at this position, or vice-versa;
  \item if some subnet action expression in the vector does not match (by
    pattern-matching) with the corresponding action expression in $C$;
  \item or if the whole set of active subnet actions in the vector
    cannot be matched (by unification) with the corresponding action expressions
      in the tuple $C$ (this is strictly stronger than the
      previous filter). 
\end {itemize}

@@

  Conversely, if $|S|_{L1_{reach}}=n+1$, we get a similar chain in
  $L1_{reach}$, with the transition $S_n \xrightarrow{l_{n}}S$.  Then $ S_n \in
  States(L1_{reach})$ and $l_n$ is satisfiable. By induction, we have
  $S_n \in States(L2_{reach})=States(L2_{sat})$, so $l_n$ will have
  been checked on the fly, in line 6 of Algorithm~\ref{alg3}. Then
  the transition $S_n \xrightarrow{l_{n}}S$ will be added to $L2_{sat}$, and thus $S \in States(L2_{reach})$.

@@

Here the behavior of the hole ``$P$'' ($a1$ in Figure~\ref{schema:unsat-ot}), after matching with the synchronisation vectors, appears as a variable
``|a1:sva\_SVA0:15:1|'', and ``$R$'' ($acc(x2)$) as ``|(ACT delta
|x:sva\_SV4:1:2|''. Line 16 requires that $a1 = \delta(x2)$ to match
the second synchronisation vector of the root pNet node, and lines
17-18 that $a1$ does not match $\delta(x1)$ for any $x1$. 
We also display  the diagnosis (``SAT'' or ``UNSAT'') generated by Z3.

@@

Figure~\ref{schema:smt-lib} shows the decomposition of the
predicate into a set of asserts, each encoding an elementary equality,
inequality, or a guard. 
The result (SAT or UNSAT) of the final \texttt{check-sat} command in
the translation () is then decoded.

@@

\paragraph{Computed open automaton.}

@@

        \medskip
        In the basic cases ($EnableState$, $EnableData$) there is no
        benefit, and even an overhead in the case of $EnableData$,
        because of the extra step of checking SAT at the end of the
        algorithm.
But as soon as there are more OTs in the system and not all
states are reachable, these tests show a significant benefit of the ``smart'' strategy, in
the considered use-cases, with 50\% of time saving for the $FailureTimer$,
over to 80\% for the composition, depending
        on the nature of the examples, but increasing significantly
        with their complexity.

@@

The formal definitions and properties of the open pNets model were
given in \cite{henrio:Forte2016}. In the current work we describe an implementation of
the model and its semantics construction, including its interaction
with the Z3 SMT engine.
The implementation has two parts: the first is an algorithm
that builds all possible combinations of synchronisations through the
pNet hierarchical structure. The result is a so-called
open automaton, whose transitions contain predicates relating
the actions of the pNet holes and controllers. Some of the open
transitions obtained at this step  may
contain predicates which do not represent any possible concrete
instantiations. 
In the second part of the tool we use the SMT solver Z3 to check the
satisfiability of the predicate in each open transition. 
To this end, we encode into Z3 the representations of the action algebra and
the predicates before submitting them to the
Z3 solver. 
%
%\hl{%
%
We have used a running example based on two possible
encodings of the Enable operator of LOTOS. The automata computed with
the algorithm can be used to prove equational properties of the
operator.
